Rclean: A Tool for Writing Cleaner, More Transparent Code

### 2020-01-28 joss/paper.Rmd
Edits discussion

@@

Conducting analyses is challenging in that it requires thinking about
multiple concepts at the same time. What did I measure? What analyses
are relevant to them? Do I need to transform the data? How do I go
about managing the data given how they were entered? What's the code
for the analysis I want to run? And so on. Data analysis can be messy
and complicated, so it's no wonder that code reflects this. And this
is a reason why having a way to isolate code based on variables can be
valuable. The following is an example of a script that has some
complications. As you can see, although the script is not extremely
long, it's long enough to make it frustrating to visualize it in its
entirety and pick through it.

@@

@@

The ``Rclean`` package provides a simple, easy to use tool for
scientists conducting analyses in the R programming language. Using
graph analytic algorithms, ``Rclean`` isolates the code necessary to
produce a specified result (e.g., an object stored in memory or a
table or figure written to disk). As statistical programming becomes
more common across the sciences, tools that make the production of
accessible code will be an important aid for improving scientific
reproducibility. ``Rclean`` has been designed to take advantage of
recent advances in data provenance capture techniques to create a
minimal tool for this purpose.

It is worth mentioning and discussing that `Rclean` does not keep
comments present in code. Although, there is often very useful or even
invaluable information in comments, the `clean` function removes
comments. This is primarily due to the lack of a mathematically formal
method for determining their relationship to the code itself. Comments
at the end of lines are typically relevant to the line they are on,
but this is not explicitly required. Also, comments occupying their
own lines usually refer to the following lines, but this is also not
necessarily the case. As `clean` depends on the unambiguous
determination of relationships in the production of results, it cannot
operate automatically on comments. However, comments in the original
code remain untouched and can be used to inform the reduced
code. Also, as the `clean` function is oriented toward isolating code
based on a specific result, the resulting code tends to naturally
support the generation of new comments that are higher level
(e.g. "The following produces a plot of the mean response of each
treatment group."), and lower level comments are not necessary because
the code is simpler and clearer.

The existing framework could be extended to support new provenance
capture methods. One possibility is *retrospective provenance*, which
tracks a computational process as it is executing. Through this
active, concurrent monitoring, retrospective provenance can gather
information that static prospective provenance can't. Only using
prospective provenance means that the outcomes of some processes can
not be predicted. For example, if there is a part of a script that is
determined by a random number, the current implementation of
prospective provenance can not predict the path that will be taken
through the code. Therefore, the code cannot be reduced to exclude the
pathway that would not be taken. However, using retrospective provenance comes
at a cost. In order to gather it, the script needs to
be executed. When scripts are computationally intensive or contain
bugs that stop execution, then retrospective provenance can not be
obtained for part or all of the code. Some work has already been done
in the direction of implementing retrospective provenance for code
cleaning in R (see http://end-to-end-provenance.github.io).

To conclude, we hope that `Rclean` makes writing scientifi software
easier. We look forward to feedback and help with extending its
applications, particularly in the area of reproducibility, such as
using code cleaning in the creation of more robust capsules
[@Pasquier2018]. To get involved, report bugs, suggest features,
please visit the project page.

>>>>

Relatedly, it is important to briefly point out that `Rclean` *does
not* keep comments present in code. This could be seen as a limitation
of the data provenance, which currently does not assign them a
relationships. Therefore, although there is often very useful or even
invaluable information in comments, the `clean` function removes
them. This is a general issue with automated methods for detecting the
relationships between comments and the code itself. Comments at the
end of lines are typically relevant to the line they are on, but this
is not a requirement and could refer to other lines. Also, comments
occupying their own lines usually refer to the following lines, but
this is also not necessarily the case. As `clean` depends on the
unambiguous determination of relationships in the production of
results, it cannot operate automatically on comments. However,
comments in the original code remain untouched and can be used to
inform the reduced code. Also, as the `clean` function is oriented
toward isolating code based on a specific result, the resulting code
tends to naturally support the generation of new comments that are
higher level (e.g. "The following produces a plot of the mean response
of each treatment group."), and lower level comments are not necessary
because the code is simpler and clearer.

@@

Conducting analyses is challenging in that it requires thinking about
multiple concepts at the same time. What did I measure? What analyses
are relevant to them? Do I need to transform the data? What's the code
for the analysis I want to run? And so on. Data analysis can be messy
and complicated, so it's no wonder that the code often reflects
this. This is why having a way to isolate code based on variables can
be valuable. The following is an example of a script that has some
complications. As you can see, although the script is not extremely
long or complicated, it's difficult enough to make it frustrating to
visualize it in its entirety and pick through it.

@@

```{R ex-keep}

```{R ex-get_vars}

```{R ex-code_graph}

```{R ex-get_libs}

@@

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring their code. Using
``Rclean`` the code necessary to produce a specified result (e.g., an
object stored in memory or a table or figure written to disk) can be
easily isolated even when tangled with code for other results. This
functionality is enabled by graph analysis of data provenance
collected from the user's script. As statistical programming becomes
more common across the sciences, tools that make the production of
accessible code will be an important aid for improving scientific
reproducibility.

Potential to integrate with a larger context of producing reproducible
projects.

- style = styler, formatr
- naming = lintr, goodpractice
- workflow = drake
- libraries = trackr, rrtools, packrat
- reproducible examples = reprex
- data = pins

Also, future work could extend the existing framework to support new
and different provenance capture methods. One possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Only using prospective provenance means
that the outcomes of some processes can not be predicted. For example,
if there is a part of a script that is determined by a random number,
the current implementation of prospective provenance can not predict
the path that will be taken through the code. Therefore, the code
cannot be reduced to exclude the pathway that would not be
taken. However, using retrospective provenance comes at a cost. In
order to gather it, the script needs to be executed. When scripts are
computationally intensive or contain bugs that stop execution, then
retrospective provenance can not be obtained for part or all of the
code. Some work has already been done in the direction of implementing
retrospective provenance for code cleaning in R (see
http://end-to-end-provenance.github.io).

To conclude, we hope that `Rclean` makes writing scientific software
easier. We look forward to feedback and help with extending its
applications, particularly in the area of reproducibility, such as
using code cleaning in the creation of more robust capsules
[@Pasquier2018]. To get involved, report bugs, suggest features,
please visit the project page.
